RUPEE: A fast and accurate purely geometric protein structure search

### 2019-01-10 docs/rupee_2.tex
lsh and min-hashing

Min-hashing is used to randomly select items from a set of items by repeatedly randomly hashing the items, sorting the hashes into a list, and then selecting the minimum item in each permuted list.
If the same random permutation of items is performed on each set of items in a collection, the key result is that the probability of matching min-hashes is equal to the Jaccard similarity \cite{Broder1998}.
In order to approximate the Jaccard similarity for a given pair of sets, a sufficient number of min-hashes must be obtained.

In our case, the items are bags of shingle hashes for protein structures from which we obtain 99 min-hashes as described in \cite{Rajaraman2012}. 
Given the key result above, the Jaccard similarity can now be approximated by the proportion of matching min-hashes.

Next, we use the LSH banding technique as described in \cite{Rajaraman2012}.
The key result of the banding technique is that if \emph{any} band positions are a match for a given pair of structures, the probability that a specific similarity threshold has been met can be calculated. 
We use 33 bands of 3-min-hashes where the probability of a Jaccard similarity of 60\% or greater is approximately 99\%. 
Banding allows the problem of finding similar items to be parallelized across bands since all that is needed for a match is a single band match.

@@

RUPEE provides two modes, fast and top-aligned. 
Each operating mode builds on the results provided by the min-hashing and LSH system described above. 
When a structure search is executed in either mode, a number of concurrent tasks are executed corresponding to the number of bands used for LSH. 
These concurrent tasks identify candidate matches based on a single band match and then validate the matches based on a comparison of min-hashes. 
In this way, RUPEE quickly compares the query structure to every structure in the database.

>>>>

Min-hashing generally is used to create a fixed-length signature for a set of items by repeatedly randomly hashing the items, sorting the hashes into a list, and then selecting the minimum hash in each permuted list.
If the same process is performed on a pair of sets, the key result is that the probability of matching min-hashes is equal to the Jaccard similarity of the two sets \cite{Broder1998}.
In order to approximate the Jaccard similarity for a given pair of sets, a sufficient number of min-hashes must be obtained.

In our case, the sets contain shingle hashes for a protein structure from which we obtain 99 min-hashes for each set as described in \cite{Rajaraman2012}. 
At this stage, we have reduced our representation of protein structures from variable-sized sets of shingles to fixed-length signatures.
Given the key result above, the Jaccard similarity for any pair of protein structures can now be approximated by the proportion of matching min-hashes.

We use the LSH banding technique as described in \cite{Rajaraman2012} to first identify candidate matches without having to compare min-hashes for every protein structure. 
We divide the 99 min-hashes in the signatures into contiguous subsequences (i.e. bands) of 3 min-hashes each, giving a total of 33 bands. 
For each band, the min-hashes it contains are combined into yet another hash. 
The key result of the banding technique is that if \emph{any} band hashes are a match for a given pair of structures, we can calculate the probability that a specific Jaccard similarity threshold has been met.
For 33 bands of 3 min-hashes, the probability of a band match occurring for a pair of structures having a Jaccard similarity of 60\% or greater is approximately 99\%. 
Banding allows the problem of finding similar structures to be parallelized across bands since all that is needed for a candidate match is a single band match.

@@

RUPEE provides two modes, fast and top-aligned. 
Each operating mode builds on the results provided by the min-hashing and LSH system described above. 
When a structure search is executed in either mode, a number of concurrent tasks are executed corresponding to the 33 bands used for LSH. 
These concurrent tasks identify candidate matches based on a single band match; for each candidate match, we estimate the Jaccard similarity using the min-hashes. 
In this way, RUPEE quickly compares the query structure to every structure in the database, which in the case of ECOD is greater than 600,000 protein domains.
